---
title: "Reproducible_document"
author: "Yuanhang Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    code_folding: show
---

# Load required libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width=8, fig.height=6)
```

```{r}
library(SCdeconR)
```

# Download required files

Data can be downloaded from zenodos [here](https://zenodo.org/records/11374072). The .rdata file contains several objects that will be used for simulation experiments, as well as evaluations on real data:

1. ref: integrated scRNA-seq reference data for breast tissue
2. phenodata: phenotype information (celltype, etc) for scRNA-seq reference data
3. genes_to_remove: genes to be removed from the scRNA-seq reference data. Those genes include Montochondrial genes, ribosomal genes and protocol differential genes. 
4. markers: marker genes for each cell type
5. gene_length: gene length information
6. bulkdata_bbd: bulkdata from BBD cohort
7. caseviewer_anno: caseviewer annotation for BBD cohort
8. bulkdata_gtex: bulkdata from gtex cohort
9. gtex_anno: pathology annotation for gtex breast cohort


```{r}
load('inputdata.rdata')
```

# Simulation studies

## Baseline

```{r}
set.seed(1234)
## use 50% of the reference data to generate artificial bulk samples
## the remaining 50% is provided during inference
idx <- sample(1:ncol(ref), ncol(ref)/2, replace = FALSE)
bulk_sim <- bulk_generator(ref = ref[,idx], 
                            phenodata = phenodata[idx,], 
                            num_mixtures = 500, 
                            num_mixtures_sprop = 10, 
                            replace = FALSE)

## perform deconvolution using different algorithms
## example provided for OLS algorithm
decon_res_baseline <- scdecon(bulk = bulk_sim$bulk, 
                      ref = ref[,-idx], 
                      phenodata = phenodata[-idx,], 
                      filter_ref = TRUE, 
                      genes_to_remove = genes_to_remove,
                      marker_genes = markers,
                      decon_method = "OLS", # specify deconvolution method 
                      norm_method_sc = "LogNormalize", # specify norm method for reference data
                      norm_method_bulk = "TPM", # specify norm method for bulk data
                      trans_method_sc = "none", # specify trans method for reference data
                      trans_method_bulk = "none", # specify trans method for bulk data
                      marker_strategy = "all",
                      gene_length = gene_length[gene_length$GeneName %in% rownames(bulk_sim$bulk),],
                      verbose = TRUE)

## generate bar plot of predicted proportions
prop_barplot(prop = decon_res_baseline[[1]], interactive = FALSE)

## Compute RMSE, bias & variance metrics 
metrics_baseline <- compute_metrics(prop_pred = decon_res_baseline[[1]], prop_sim = bulk_sim$prop)
```

## Cell type removal 

```{r}
## perform deconvolution using different algorithms
## example provided for OLS algorithm
## need to specify to_remove parameter
decon_res_ct_rm <- scdecon(bulk = bulk_sim$bulk, 
                      ref = ref[,-idx], 
                      phenodata = phenodata[-idx,], 
                      filter_ref = TRUE, 
                      genes_to_remove = genes_to_remove,
                      marker_genes = markers,
                      decon_method = "OLS", # specify deconvolution method 
                      norm_method_sc = "LogNormalize", # specify norm method for reference data
                      norm_method_bulk = "TPM", # specify norm method for bulk data
                      trans_method_sc = "none", # specify trans method for reference data
                      trans_method_bulk = "none", # specify trans method for bulk data
                      marker_strategy = "all",
                      gene_length = gene_length[gene_length$GeneName %in% rownames(bulk_sim$bulk),],
                      to_remove = "Adipocyte", # specify cell type to remove
                      prop = bulk_sim$prop, # specify prop matrix
                      verbose = TRUE)

## generate bar plot of predicted proportions
prop_barplot(prop = decon_res_ct_rm[[1]], interactive = FALSE)

## Compute RMSE, bias & variance metrics 
metrics_ct_rm <- compute_metrics(prop_pred = decon_res_ct_rm[[1]], prop_sim = decon_res_ct_rm[[3]])
```

## FFPE

```{r}
## perform deconvolution using different algorithms
## example provided for OLS algorithm
## need to enable ffpe_artifacts parameter
## download model object from: https://github.com/Liuy12/SCdeconR/blob/master/data/ffpemodel.rda
data("ffpemodel")
decon_res_ffpe <- scdecon(bulk = bulk_sim$bulk, 
                      ref = ref[,-idx], 
                      phenodata = phenodata[-idx,], 
                      filter_ref = TRUE, 
                      genes_to_remove = genes_to_remove,
                      marker_genes = markers,
                      decon_method = "OLS", # specify deconvolution method 
                      norm_method_sc = "LogNormalize", # specify norm method for reference data
                      norm_method_bulk = "TPM", # specify norm method for bulk data
                      trans_method_sc = "none", # specify trans method for reference data
                      trans_method_bulk = "none", # specify trans method for bulk data
                      marker_strategy = "all",
                      gene_length = gene_length[gene_length$GeneName %in% rownames(bulk_sim$bulk),], 
                     ffpe_artifacts = TRUE, # enable FFPE artifact
                     model = ffpemodel,
                      verbose = TRUE)

## generate bar plot of predicted proportions
prop_barplot(prop = decon_res_ffpe[[1]], interactive = FALSE)

## Compute RMSE, bias & variance metrics 
metrics_ffpe <- compute_metrics(prop_pred = decon_res_ffpe[[1]], prop_sim = bulk_sim$prop)
```

# Benchmark using BBD FFPE samples

```{r}
## perform deconvolution using different algorithms
## example provided for OLS algorithm
## using all data from ref
decon_res_bbd <- scdecon(bulk = bulkdata_bbd, 
                      ref = ref, 
                      phenodata = phenodata, 
                      filter_ref = TRUE, 
                      genes_to_remove = genes_to_remove,
                      marker_genes = markers,
                     min_pct_ct = 0.2, # specify minimum proportion of expressing cells
                      decon_method = "OLS", # specify deconvolution method 
                      norm_method_sc = "LogNormalize", # specify norm method for reference data
                      norm_method_bulk = "TPM", # specify norm method for bulk data
                      trans_method_sc = "none", # specify trans method for reference data
                      trans_method_bulk = "none", # specify trans method for bulk data
                     marker_strategy = "all",
                      gene_length = gene_length[gene_length$GeneName %in% rownames(bulkdata_bbd),],
                      verbose = TRUE)

## generate bar plot of predicted proportions
prop_barplot(prop = decon_res_bbd[[1]], interactive = FALSE)

## Compute RMSE, bias & variance metrics 
## extract predicted proportions for selected cell types
prop_pred <- decon_res_bbd[[1]][c("Epithelial_cells", "Fibroblast", "Endothelial_cells", "Pericytes", "Adipocyte"),]
prop_pred <- rbind(prop_pred, apply(prop_pred[2:4,], 2, sum))
prop_pred <- prop_pred[-c(2:4),]
rownames(prop_pred)[3] <- "Stroma"
prop_pred <- sweep(prop_pred, 2, apply(prop_pred, 2, sum), "/")
metrics_ffpe <- compute_metrics(prop_pred = prop_pred, prop_sim = caseviewer_anno)
## as there is no technical replicates in BBD cohort, the first element of metrics_ffpe probably 
## don't make much sense. The second element represents samples-level RMSEs.
```

# Evaluation using GTex breast tissue cohort

```{r}
## example provided for OLS algorithm
## Original paper used scaden for prediction
## using all data from ref
decon_res_gtex <- scdecon(bulk = bulkdata_gtex, 
                      ref = ref, 
                      phenodata = phenodata, 
                      filter_ref = TRUE, 
                      genes_to_remove = genes_to_remove,
                      min_pct_ct = 0.2, # specify minimum proportion of expressing cells
                      marker_genes = markers,
                      decon_method = "OLS", # specify deconvolution method 
                      norm_method_sc = "LogNormalize", # specify norm method for reference data
                      norm_method_bulk = "TPM", # specify norm method for bulk data
                      trans_method_sc = "none", # specify trans method for reference data
                      trans_method_bulk = "none", # specify trans method for bulk data
                      marker_strategy = "all",
                      gene_length = gene_length[gene_length$GeneName %in% rownames(bulkdata_gtex),],
                      verbose = TRUE)

## generate bar plot of predicted proportions
prop_barplot(prop = decon_res_gtex[[1]], interactive = FALSE)
```
